Generation of Variation and Mean Fitness Increase: \\ Necessity is the Mother of Genetic Invention

### 2017-11-28 ms.tex
cite Lukacisinova2017

\paragraph*{General model.}\label{sec:general_model}
Consider a large  population with an arbitrary set of types $A_1, A_2, \ldots, A_n$.
The frequency and fitness of individuals of type $A_k$ are $f_k$ and $w_k$, respectively.
The probability that an individual of type $A_k$ will transition to some other type is $C_k$, and given a transition occurs, the probability that it will transition to type $A_j$ is $M_{j,k}$.
Therefore, the change in the frequencies of type $A_k$ is described by the transformation $f \to f'$: 
\begin{equation}
\bar{w} f'_k = (1-C_k) w_k f_k + \sum_{j=1}^{n}{C_j M_{k,j} w_j f_j},
\label{eq:model_sum}
\end{equation}
or in matrix form
\begin{equation}
\bar{w} f' = (\ci - C + MC)D f,
\label{eq:model}
\end{equation}
where $f=(f_1, f_2, \ldots, f_n)$ is a frequency vector with $f_k\ge 0$ and  $\sum_{k=1}^n{f_k} = 1$;
$D$ is a positive diagonal matrix with  elements $w_k$ such that $w_k \ne w_j$ for some $k \ne j$; 
%and we can assume $0 < w_k \le 1$, i.e., fitness values are positive and relative to the fitness of the fittest type; 
$C$ is a positive diagonal  matrix with  elements $C_k$;
$M$ is a primitive column-stochastic matrix: $M_{j,k} \ge 0$ for all $j,k$, $\sum_{j=1}^n {M_{j,k}} = 1$ for all $k$, and $(M^{\ell})_{j,k} > 0$ for all $j,k$ for some positive integer $\ell$;
$\ci$ is the $n\times n$ identity matrix;
and $\bar{w}$ is the normalizing factor such that $\sum_{k=1}^n{f'_k}=1$ and is equal to the population mean fitness $\bar w=\sum_{k=1}^n f_kw_k$.

@@

The equilibrium of Eqs.~\ref{eq:modifier_model} when modifier allele $M$ is absent from the population is $(\hat v,0)$, where $\hat v$ is given in Eq.~\ref{eq:model_equilibrium} and $g_k=0$ for all $k$. 
The stability of allele $m$ to invasion by allele $M$ is determined by the leading eigenvalue $\lambda_1$ of $\cl_{ex}$ the external stability matrix of the equilibrium $(\hat v,0)$, which, in turn, is determined by the Jacobian $\cj$ of the system in Eqs.~\ref{eq:modifier_model} evaluated at the equilibrium $(\hat v,0)$, where
\begin{equation}
\cj = \begin{pmatrix}\cl_{in} & 0 \\ 0 & \cl_{ex} \end{pmatrix},
\end{equation}
and $\cl_{in}$ is the local stability matrix of the equilibrium $(\hat v,0)$ in the space $\sum_{k=1}^n{v_k}=1$.
The zero block matrices are due to the complete linkage between the modifier and the types $A_k$ and to the lack of transition (i.e., mutation) between the modifier alleles.

$\cl_{ex}$ can be written as
\begin{equation}\label{eq:Lex}
\cl_{ex} = (I - \tilde{C} + M \tilde{C}) \tilde{D} ,
\end{equation}
where $\tilde{D} =D/\hat{\bar{w}}$ is the diagonal matrix with elements $w_k/\hat{\bar{w}}$ for all $k$, and $\hat{\bar{w}}$ is the stable population mean fitness in the absence of the modifier allele $M$.
$\lambda_1$, the leading eigenvalue of $\cl_{ex}$, coincides with $\tilde{\bar{w}}$, the maximal mean fitness associated with (28).
Thus we can apply Result 1 to $\lambda_1$ and obtain the following result. %TODO refs, maximal?

@@

Note that if the modifier has the same effect on all types,
then we can substitute $C = \mu I$ (with $\mu > 0$) in Eq.~\ref{eq:model_equilibrium}, and proceeding as in Eq.~\ref{eq:mfp_K}, we find a relationship previously described by~\citet[eq.~24]{Hermisson2002},
\begin{equation}\label{eq:mutational_loss}
\frac{\partial \hat{\bar{w}}}{\partial \mu} = 
-\frac{1}{\mu}\Big(\sum_{k=1}^{n} \hat u_k \hat v_k w_k - \hat{\bar{w}} \Big) = -\frac{1}{\mu} G,
\end{equation}
where $G$, equal to the difference between the \emph{ancestral mean fitness} $(\sum_{k=1}^{n} \hat u_k \hat v_k w_k)$ and the \emph{stable population mean fitness} $\hat{\bar{w}}$ when $C_k=\mu$, is  called the \emph{mutational loss}~\citep{Hermisson2002}.

If an invading  allele $M$ reduces the transition probability compared to the resident allele $m$, i.e. $C_k = \mu$ and $\tilde{C}_k = \tilde\mu$ with $\tilde\mu<\mu$,  then $\cl_{ex}$ becomes
\begin{equation}
\cl_{ex} = \bigl[(1 - \tilde\mu)I + \tilde\mu M\bigr] \bar D,
\end{equation} 
where $\bar D =D/\bar w$ is  the diagonal matrix with entries $w_k/\bar w$ for $k=1,2,\dots,n$. Using the unified reduction principle \citep{Altenberg2017} we know that the leading eigenvalue $\lambda_1$  of $\cl_{ex}$ in Eq.\ (35) satisfies $\partial\lambda_1/\partial\tilde\mu<0$. Thus we can conclude that the mutational loss $G$ is positive and the ancestral mean fitness is higher than the population mean fitness. 
% TODO check above for refs, hats, etc.

@@

Our analysis focuses on populations at equilibrium.
Nevertheless, it has been demonstrated that during adaptive evolution (i.e., in non-equilibrium populations), a modifier that increases the mutation rate of maladapted individuals can be favored by selection~\citep{Ram2012} and increase the adaptation rate~\citep{Ram2014},
and empirical evidence suggests that \emph{stress-induced mutagenesis} is common in bacteria and yeast, and may be prevalent in plants, flies, and human cancer cells~\citep{Rosenberg2012,Fitzgerald2017b}.
Similar theoretical results have been demonstrated for a modifier that increases the recombination rate in maladapted individuals~\citep{Hadany2003b,Hadany2003a}.

@@

\citet{Caswell1978} gave a \emph{formula for the sensitivity of the population growth rate to changes in life history parameters}.
In this formula, the \emph{population growth rate} is the leading eigenvalue of the population transformation matrix $T$, the \emph{life history parameters} are elements of $T$, and the \emph{sensitivity} is the derivative of the former with respect to the latter.
This is a useful formula~\citep[ch.~10]{Caswell1978,Hermisson2002,Ram2012,Otto2007}, and therefore we reproduce it here.

@@

% Appendix: Fisher's reproductive value
\subsection*{Appendix B}\label{sec:AppB}
\begin{remark}[Fisher's reproductive value]
Let $M$ be an irreducible column-stochastic matrix and $D$ be a positive diagonal matrix. 
The elements of the left \emph{Perron} eigenvector $\hat u$ of the matrix $MD$ can be regarded as \emph{Fisher's reproductive values}~\citep[pg.~27]{Fisher1930}
\end{remark}

>>>>

\paragraph*{General model.}\label{sec:general_model}
Consider a large  population with an arbitrary set of types $A_1, A_2, \ldots, A_n$.
The frequency and fitness of individuals of type $A_k$ are $f_k$ and $w_k$, respectively.
The probability that an individual of type $A_k$ will transition to some other type is $C_k$, and given a transition occurs, the probability that it will transition to type $A_j$ is $M_{j,k}$.
Therefore, the change in the frequencies of type $A_k$ is described by the transformation $f \to f'$: 
\begin{equation}
\bar{w} f'_k = (1-C_k) w_k f_k + \sum_{j=1}^{n}{C_j M_{k,j} w_j f_j},
\label{eq:model_sum}
\end{equation}
or in matrix form
\begin{equation}
\bar{w} f' = (\ci - C + MC)D f,
\label{eq:model}
\end{equation}
where $f=(f_1, f_2, \ldots, f_n)$ is a frequency vector with $f_k\ge 0$ and  $\sum_{k=1}^n{f_k} = 1$;
$D$ is a positive diagonal matrix with entries $w_k$ such that $w_k \ne w_j$ for some $k \ne j$; 
$C$ is a positive diagonal  matrix with entries $C_k$;
$M$ is a primitive column-stochastic matrix: $M_{j,k} \ge 0$ for all $j,k$, $\sum_{j=1}^n {M_{j,k}} = 1$ for all $k$, and $(M^{\ell})_{j,k} > 0$ for all $j,k$ for some positive integer $\ell$;
$\ci$ is the $n\times n$ identity matrix;
and $\bar{w}$ is the normalizing factor such that $\sum_{k=1}^n{f'_k}=1$ and is equal to the population mean fitness $\bar w=\sum_{k=1}^n f_kw_k$.

@@

The equilibrium of Eqs.~\ref{eq:modifier_model} when modifier allele $M$ is absent from the population is $(\hat v,0)$, where $\hat v$ is given in Eq.~\ref{eq:model_equilibrium} and $g_k=0$ for all $k$. 
The stability of allele $m$ to invasion by allele $M$ is determined by the leading eigenvalue $\lambda_1$ of $\cl_{ex}$ the external stability matrix of the equilibrium $(\hat v,0)$, which, in turn, is determined by the Jacobian $\cj$ of the system in Eqs.~\ref{eq:modifier_model} evaluated at the equilibrium $(\hat v,0)$, where
\begin{equation}
\cj = \begin{pmatrix}\cl_{in} & 0 \\ 0 & \cl_{ex} \end{pmatrix},
\end{equation}
and $\cl_{in}$ is the local stability matrix of the equilibrium $(\hat v,0)$ in the space $\sum_{k=1}^n{f_k}=1$.
The zero block matrices are due to the complete linkage between the modifier and the types $A_k$ and to the lack of transition (i.e., mutation) between the modifier alleles.

$\cl_{ex}$ can be written as
\begin{equation}\label{eq:Lex}
\cl_{ex} = (I - \tilde{C} + M \tilde{C}) \tilde{D} ,
\end{equation}
where $\tilde{D} =D/\hat{\bar{w}}$ is the diagonal matrix with entries $w_k/\hat{\bar{w}}$ for all $k$, and $\hat{\bar{w}}$ is the stable population mean fitness in the absence of the modifier allele $M$.
$\lambda_1$, the leading eigenvalue of $\cl_{ex}$, coincides with $\tilde{\bar{w}}$, the maximal mean fitness associated with Eq.~\ref{eq:Lex}.
Thus we can apply Result~\ref{result:mfp} to $\lambda_1$ and obtain the following result.

@@

Note that if the modifier has the same effect on all types,
then we can substitute $C = \mu I$ (with $\mu > 0$) in Eq.~\ref{eq:model_equilibrium}, and proceeding as in Eq.~\ref{eq:mfp_K}, we find a relationship previously described by~\citet[eq.~24]{Hermisson2002},
\begin{equation}\label{eq:mutational_loss}
\frac{\partial \hat{\bar{w}}}{\partial \mu} = 
-\frac{1}{\mu}\Big(\sum_{k=1}^{n} \hat u_k \hat v_k w_k - \hat{\bar{w}} \Big) = -\frac{1}{\mu} G,
\end{equation}
where $G$, equal to the difference between the \emph{ancestral mean fitness} $(\sum_{k=1}^{n} \hat u_k \hat v_k w_k)$ and the \emph{stable population mean fitness} $(\hat{\bar{w}})$ when $C_k=\mu$, is  called the \emph{mutational loss}~\citep{Hermisson2002}.

If an invading  allele $M$ reduces the transition probability compared to the resident allele $m$, i.e., $C_k = \mu$ and $\tilde{C}_k = \tilde{\mu}$ for all $k$ with $\tilde{\mu}<\mu$, then $\cl_{ex}$ becomes
\begin{equation} \label{eq:Lex_mu}
\cl_{ex} = \bigl[(1 - \tilde\mu)I + \tilde\mu M\bigr] \tilde{D},
\end{equation} 
where $\tilde{D} =D/\hat{\bar{w}}$ is  the diagonal matrix with entries $w_k/\hat{\bar{w}}$ for all $k$.
Using the \emph{unified reduction principle}~\citep{Altenberg2017} we know that the leading eigenvalue $\lambda_1$  of $\cl_{ex}$ in Eq.~\ref{eq:Lex_mu} satisfies $\partial\lambda_1 / \partial\tilde{\mu} < 0$.
Thus we can conclude that the mutational loss $G$ is positive and the ancestral mean fitness is higher than the stable population mean fitness.

@@

Our analysis focuses on populations at equilibrium.
Nevertheless, it has been demonstrated that during adaptive evolution (i.e., in non-equilibrium populations), a modifier that increases the mutation rate of maladapted individuals can be favored by selection~\citep{Ram2012,Lukacisinova2017} and increase the adaptation rate~\citep{Ram2014},
and empirical evidence suggests that \emph{stress-induced mutagenesis} is common in bacteria and yeast, and may be prevalent in plants, flies, and human cancer cells~\citep{Rosenberg2012,Fitzgerald2017b}.
Similar theoretical results have been demonstrated for a modifier that increases the recombination rate in maladapted individuals~\citep{Hadany2003b,Hadany2003a}.

@@

\citet{Caswell1978} gave a \emph{formula for the sensitivity of the population growth rate to changes in life history parameters}.
In this formula, the \emph{population growth rate} is the leading eigenvalue of the population transformation matrix $T$, the \emph{life history parameters} are entries of $T$, and the \emph{sensitivity} is the derivative of the former with respect to the latter.
This is a useful formula~\citep[ch.~10]{Caswell1978,Hermisson2002,Ram2012,Otto2007}, and therefore we reproduce it here.

@@

% Appendix: Fisher's reproductive value
\subsection*{Appendix B}\label{sec:AppB}
\begin{remark}[Fisher's reproductive value]
Let $M$ be an irreducible column-stochastic matrix and $D$ be a positive diagonal matrix. 
The entries of the left \emph{Perron} eigenvector $\hat u$ of the matrix $MD$ can be regarded as \emph{Fisher's reproductive values}~\citep[pg.~27]{Fisher1930}
\end{remark}
